---
title: "PSTAT 131 Homework 1"
author: "Jennifer Rink"
date: "2023-01-20"
output:
  rmarkdown::html_document:
    toc: true
    toc_float: true
    theme: cerulean
    highlight: espresso
    smooth_scroll: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

---

## Machine Learning Main Ideas ##

### Question 1 ###
Define supervised and unsupervised learning. What are the difference(s) between them?  
  
**Solution**:  
Supervised learning involves building a statistical model that predicts/estimates an output based on one or more inputs.  
Unsupervised learning involves building a statistical model that includes inputs but doesn't have a supervising output. [1, ISLR]
  
  
  
### Question 2 ###
Explain the difference between a regression model and a classification model, specifically in the context of machine learning.  
  
**Solution**:  
The Regression Model predicts a **continuous** (numerical) output value, Y.  
The Classification Model predicts a **categorical** (non-numerical) output value, Y. [3, ISLR]
  
  
  
### Question 3 ###
Name two commonly used metrics for regression ML problems. Name two commonly used metrics for classification ML problems. 
  
**Solution**:  
Two common metrics for Regression ML: **Mean Squared Error and Root Mean Squared Error**.  
Two common metrics for Classification ML: **Error Rate and Area Under ROC Curve**. [From Lecture Week 1]
  
  
  
### Question 4 ###
As discussed, statistical models can be used for different purposes. These purposes can generally be classified into the following three categories. Provide a brief description of each.  
  
**Solution**:   
Descriptive models: Used to visually emphasize a trend in data.

Inferential models: Used to test theories, emphasize significant features, and state the relationship between outcome and predictors.

Predictive models: Used to predict Y with minimum reducible error, not focused on hypothesis tests. [From Lecture Week 1]
  
  
  
### Question 5 ###
Predictive models are frequently used in machine learning, and they can usually be described as either mechanistic or empirically-driven. Answer the following questions.

Define mechanistic. Define empirically-driven. How do these model types differ? How are they similar?

In general, is a mechanistic or empirically-driven model easier to understand? Explain your choice.

Describe how the bias-variance tradeoff is related to the use of mechanistic or empirically-driven models.  
  
**Solution**:   

  A **mechanistic** model uses theory to predict an outcome.  
  A **empirical** model uses data to develop a theory.  
    [Both models are flexible]  
  
  I think mechanistic modeling is easier to understand because it is easier to access real life data to make predictions. With mechanistic modeling, we build up a model from a basic understanding of what might be the outcome in real life. It is a good place to start when doing regression and attempting to fit your own model.  
      
  Both mechanistic and empirical models can be the victim of overfitting which is related to the Bias-Variance Trade-off theorem. The theorem describes a trade-off between the flexibility of the model and how accurately it predicts. To build a good model, we want to find a balance between the bias and variance such that the total error is minimized.  
  
  
### Question 6 ###
A political candidate’s campaign has collected some detailed voter history data from their constituents. The campaign is interested in two questions:

Given a voter’s profile/data, how likely is it that they will vote in favor of the candidate?

How would a voter’s likelihood of support for the candidate change if they had personal contact with the candidate?

Classify each question as either predictive or inferential. Explain your reasoning for each.  
  
**Solution**:   
  The first question is inferential because you are given data before predicting the outcome.
  The second question is predictive because it is learning a pattern in the past to predict the future outcome. 
  
---
  
## Exploratory Data Analysis ##
  
### Exercise 1 ###
We are interested in highway miles per gallon, or the hwy variable. Create a histogram of this variable. Describe what you see/learn.    
  
**Solution**:   
```{r}
library(ggplot2)  # load in necessary libraries
library(tidyverse)

hist(mpg$hwy, col="red")
```
  
  This histogram shows that the HWY variable is bimodal and that the most frequently observed values for highway miles per gallon is between 25mpg-30mpg and the second most frequently observed values is between 15mpg-20mp.
  
### Exercise 2 ###
Create a scatterplot. Put hwy on the x-axis and cty on the y-axis. Describe what you notice. Is there a relationship between hwy and cty? What does this mean?    
  
**Solution**:   
```{r}
ggplot(mpg, aes(x=hwy, y=cty), main="Scatterplot Highway VS City Mileage", xlab="Highway Mileage", ylab="City Mileage" ) + geom_point()
```
  
There does seem to be a positive linear relationship between Highway Mileage and City Mileage; which means that as the value of one increases, the other's value increases as well. This makes sense because the relationship of gas mileage efficiency should be correlated. 
  
    
### Exercise 3 ###
Make a bar plot of manufacturer. Flip it so that the manufacturers are on the y-axis. Order the bars by height. Which manufacturer produced the most cars? Which produced the least?    
  
**Solution**:   
```{r}
mpg_manu<-table(mpg$manufacturer)
barplot(sort(mpg_manu, decreasing=FALSE), horiz=TRUE, las=2, cex.names = 0.7)
```
  
Dodge manufactured the most cars and Lincoln manufactured the least.
    
### Exercise 4 ###
Make a box plot of hwy, grouped by cyl. Use geom_jitter() and the alpha argument to add points to the plot.
Describe what you see. Is there a relationship between hwy and cyl? What do you notice?
    
  
**Solution**:   
```{r}
plot_box<-ggplot(mpg, aes(group=cyl, x=cyl, y=hwy))
plot_box+geom_boxplot()+geom_jitter(alpha=0.5)
```
  
The Boxplots for cyl(4) and cyl(5) have the same median but the HWY values for cyl(4) are much more variable than cyl(5). 
  
Cyl(4) and cyl(8) both have a few outliers.
  
The relationship I noticed between HWY and CYL is that the smaller the number of cylinders, the higher the mileage on the highway is; and therefore the more cylinders in a car, the lower the mileage on the highway is. (They have a negative relationship)
  
### Exercise 5 ###
Use the corrplot package to make a lower triangle correlation matrix of the mpg dataset. (Hint: You can find information on the package here.)

Which variables are positively or negatively correlated with which others? Do these relationships make sense to you? Are there any that surprise you?    
  
**Solution**:   
```{r}
# First, remove categorical variables from the dataset
no_categorical_var_mpg<-select(mpg, displ, year, cyl, cty, hwy)

mat<-cor(no_categorical_var_mpg)
corrplot::corrplot(mat, type="lower", method="number")
```  
  
(HWY and DISPL), (CTY and DISPL), (HWY and CYL), and (CTY and CYL) are strongly negatively correlated.
  
(CYL and DISPL) and (HWY and CTY) are strongly positively correlated.
  
(YEAR and DISPL) and (CYL and YEAR) are very weakly positively correlated. 
  
(CTY and YEAR) are extremely weakly negatively correlated.
  
  
The only relationship I'm a bit confused about is between (CTY and DISPL) and (HWY and DISPL). I would think that the more an engine displaces the farther it would go, meaning it would have a positive relationship with both CTY and HWY.
